A significance test in forward stepwise regression

### 2014-04-04 paper/paper.tex
working on quadratic selection

% put your definitions there:
\startlocaldefs
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphics,amssymb,color}
\usepackage{graphicx,subfigure}
%\usepackage[boxed]{algorithm2e}

@@

In the ordinary least squares setting, a significance test for a
single variable can be conducted by comparing the drop in residual
sums of squares (RSS) to a $\chi^2_1$ distribution. Similarly, when
adding a group of $k$ variables we can compare the drop in RSS to a
$\chi^2_k$ random variable. This generally does not work when the
group to be added has been chosen by a method that uses the data,
and in particular it fails for forward stepwise procedures adding
the ``best'' (e.g. most highly correlated) predictor in each step. In
that case, the test statistic (drop in RSS) does not match the
theoretical null distribution even when the null hypothesis is
true. \cite{significance:lasso} introduced a new test statistic based
on the knots in the lasso solution path. They derived a simple
asymptotic null distribution, proved a type of convergence under broad
``minimum growth'' conditions, and demonstrated in simulations that
the test statistic closely matches its asymptotic distribution even in
finite samples.  That work marked an important advance in the problem
of combining inference with model selection.  \cite{tests:adaptive}
extended that work to the group lasso \citep{grouplasso} and other
problems, and modified the test statistic to one with an exact finite
sample distribution under the global null hypothesis.

@@

In the setting of grouped variables, we use an analogous test statistic to
the right hand side of \cite{test:exact}. The test statistic is presented as an example
in \cite{tests:adaptive} though we re-derive it here in simpler form. The test draws inspiration
from the group LASSO.

@@

\todo Fix references below

The group lasso estimator is discussed in (ref that guy's thesis) and
(ref Yuan and Lin).  An important extension is the sparse group lasso
(ref ???) which enforces sparsity in groups as well as sparsity of the
coefficients within the groups.  For a survey on group lasso and
related factor models see (ref???). \todo review some more literature
and add a few more references here if they seem worthwhile.

\begin{figure}
\begin{center}
\resizebox{!}{3in}{\input{curved.pdf_tex}}
\end{center}
Selection conditioned on an arbitrary selection event.
\end{figure}
We are interested in inference after a selection procedure, say, $S$, for
which the selection event is the intersection of a list of quadratic inequalities:

$$
\begin{aligned}
\{S(y)=s\} = \cap_{i \in I(s)} \{y: y^TQ_iy + a_i^Ty\leq b_i \}.
\end{aligned}
$$

The quadratic forms are not assumed non-negative definite, but we can, without loss of generality
assume they are symmetric.

If we consider forward stepwise with groups the selection procedure is just
choosing the first group, $g^*$ and we have
$$
\{g^*(y)=g\} =  \cap_{h \neq g} \{y: y^T(X_hX_h^T/w_h^2 - X_gX_g^T/w_g^2)y \leq 0\}.
$$

We want to slice through the selection event along a ray with direction $\eta$ that passes through $y$. 
That is, we need to find
$$
\begin{aligned}
\left\{t: S(y+t\eta)=s \right\} &= \cap_{i \in I(s)} \left\{t: (y+t\eta)^T Q_i (y+t\eta) + a_i^T(y+t\eta) \leq b_i\right\}. \\
\end{aligned}
$$

For any given $i \in I(s)$ we see that
$$
\begin{aligned}
\left\{t: (y+t\eta)^T Q_i (y+t\eta) + a_i^T(y+t\eta) \leq b_i\right\} &= \left\{t: t^2 \cdot \eta^TQ_i\eta + t \cdot(2 y^TQ_i\eta + a_i^T\eta) + y^TQ_iy + a_i^Ty - b_i \leq 0 \right\} \\
&= \text{Intervals}(Q_i,a_i,b_i,y,\eta)
\end{aligned}
$$

Each value of $\text{Intervals}(Q_i,a_i,b_i,y,\eta)$ above is one of:
- $\emptyset$, in which case the selection event is also the $\emptyset$ (so this never occurs);
- $(-\infty, L(Q_i,a_i,b_i,y,\eta)] \cap [U(Q_i,a_i,b_i,y,\eta), \infty)$ if $\eta^TQ_i\eta < 0$;
- $[L(Q_i,a_i,b_i,y,\eta), U(Q_i,a_i,b_i,y,\eta)]$ if $\eta^TQ_i\eta > 0$;
- $(-\infty, L(Q_i,a_i,b_i,y,\eta)]$ if $\eta^TQ_i\eta=0$ and $2 y^TQ_i\eta + a_i^T\eta > 0$;
- $[U(Q_i,a_i,b_i,y,\eta), \infty)$ if $\eta^TQ_i\eta=0$ and $2 y^TQ_i\eta + a_i^T\eta < 0$;
- $(-\infty,\infty)$ if $\eta^TQ_i\eta=0, 2 y^TQ_i\eta + a_i^T\eta = 0$ and $y^TQ_iy + a_i^Ty - b_i \leq 0$.

@@

Finally, most of the plots below show the following information. The
horizontal axis is the step index for forward stepwise. The dashed
line shows the proportion of iterations where a truly nonzero variable
was added to the model at that step. Solid verticle lines show the
middle 50\% of p-values calculated at that step, with red calculated
on a null model with no nonzero groups, blue calculated on the non-null
model, and green calculated on the non-null model from the usual
$\chi^2$ significance test.
The points in these lines are the average p-value at that step, and
the triangles above and below show the 95\% and 5\% quantiles.

>>>>

% put your definitions there:
\startlocaldefs
\usepackage{algorithm}
\usepackage{algorithmic,amssymb}
\usepackage{graphicx,xcolor}
\usepackage{subfigure}
%\usepackage[boxed]{algorithm2e}

@@

In the ordinary least squares setting, a significance test for a
single variable can be conducted by comparing the drop in residual
sums of squares (RSS) to a $\chi^2_1$ distribution. Similarly, when
adding a group of $k$ variables we can compare the drop in RSS to a
$\chi^2_k$ random variable. This generally does not work when the
group to be added has been chosen by a method that uses the data \cite{olshen:flevel},
and in particular it fails for forward stepwise procedures adding
the ``best'' (e.g. most highly correlated) predictor in each step. In
that case, the test statistic (drop in RSS) does not match the
theoretical null distribution even when the null hypothesis is
true. \cite{significance:lasso} introduced a new test statistic based
on the knots in the lasso solution path. They derived a simple
asymptotic null distribution, proved a type of convergence under broad
``minimum growth'' conditions, and demonstrated in simulations that
the test statistic closely matches its asymptotic distribution even in
finite samples.  That work marked an important advance in the problem
of combining inference with model selection.  \cite{tests:adaptive}
extended that work to the group lasso \citep{grouplasso} and other
problems, and modified the test statistic to one with an exact finite
sample distribution under the global null hypothesis.

@@

In the setting of grouped variables, we use an analogous test statistic to
the right hand side of \eqref{test:exact}. The test statistic is presented as an example
in \cite{tests:adaptive} though we re-derive it here in simpler form. The test draws inspiration
from the group LASSO.

@@

\begin{figure}
\label{fig:polytope}
\begin{center}
\resizebox{!}{3in}{\input{../figs/polytope.pdf_tex}}
\end{center}
A polytope selection event. Solving the LASSO \eqref{eq:lasso} for a fixed value
of $\lambda$ yields both an active set $\hat{E}(y)$ and a collection of signs of 
active variables $z_{\hat{E}(y)}$. The event $\{y: (\hat{E}(y), z_{\hat{E}(y)}) = (E, z_E)\}$
can be written as a set of affine constraints.
\end{figure}

\subsection{Selection events}

The approach to post-selection inference described \cite{lasso:fixed}
describes the selection event for the LASSO as a convex polytope stylized
in Figure \ref{fig:polytope}. If all of our groups were of size 1, then
the event that we observe a given set of variables
in the forward stepwise path (with their signs as they enter) would also
be described by a set of affine inequalities. This is described in more detail in \cite{exact:lars}.

After selection, exact inference for a (randomly) chosen contrast $\eta(y)^T\mu$ of the mean vector
$\mu$ is carried by considering the conditional distribution of 
\begin{equation}
y \sim N(\mu, \sigma^2 I) \bigl| Ay \leq b.
\end{equation}
See \cite{lasso:fixed,exact:lars} for further details.

In the grouped variables setting, however, the event that $g^*(y)$, the first
group chosen, is equal to some fixed group $g$ is generally not described by
a set of affine constraints. The event can be described as
\begin{equation}
\label{eq:first:quadratic}
\begin{aligned}
\{g^*(y)=g\} &= \{ \|X_g^Ty\|_2 /w_g \geq \|X_h^Ty\|_2 /w_h , \ \forall h \neq g\} \\
&=  \{ y^T(X_g^TX_g)y / w_g - y^T(X_h^TX_h)y / w_h, \ \forall h \neq g\} \\
\end{aligned}
\end{equation}

Hence, our selection event can be expressed as 
the intersection of a list of quadratic inequalities. This non-affine selection event is stylized in Figure
\ref{fig:curved}.
In this section, we consider an arbitrary selection procedure
$S$ that returns one of a set of possible outcomes $s \in {\cal S}$ by a set of quadratic inequalities.
That is, for all possible outcomes $s \in {\cal S}$:
\begin{equation}
\label{eq:selection}
\begin{aligned}
\{S(y)=s\} = \cap_{i \neq I(s)} \{y: y^TQ_iy + a_i^Ty\leq b_i \}.
\end{aligned}
\end{equation}

The quadratic forms are not assumed non-negative definite, but we can, without loss of generality
assume they are symmetric.
As in \cite{lasso:exact} we will be interested in some randomly chosen contrast
$\eta(S(y))^T\mu$. In our grouped selection procedure, the procedure $S(y)=g^*(y)$
and $\eta$ is typically 
$$
\eta(S(y)) = \eta(g^*(y)) = \frac{X_{g^*(y)}^Ty}{\|X_{g^*(y)}^Ty\|_2}.
$$

Having fixed $\eta$, we must slice through the selection event along a ray with direction $\eta$ that passes through $y$. 
That is, we need to find
$$
\begin{aligned}
\left\{t: S(y+t \cdot \eta)=s \right\} &= \cap_{i \in I(s)} \left\{t: (y+t \cdot \eta)^T Q_i (y+t \cdot \eta) + a_i^T(y+t \cdot \eta) \leq b_i\right\}. \\
\end{aligned}
$$

\begin{figure}
\label{fig:curved}
\begin{center}
\resizebox{!}{3in}{\input{../figs/curved.pdf_edit_tex}}
\end{center}
A generic selection event, given by the intersection of a list
of quadratic inequalities. \end{figure}

The slice for any particular inequality is
$$
\begin{aligned}
\lefteqn{
\left\{t: (y+t\eta)^T Q_i (y+t \cdot \eta) + a_i^T(y+t \cdot \eta) \leq b_i\right\}} \\
 & \qquad = \left\{t: t^2 \cdot \eta^TQ_i\eta + t \cdot(2 y^TQ_i\eta + a_i^T\eta) + y^TQ_iy + a_i^Ty - b_i \leq 0 \right\} \\
\end{aligned}
$$
This can be computed easily, and results in either the empty, a single interval (possibly infinite) or the union of 
two disjoint infinite intervals. The intersection over all $I(s)$ is therefore easily computable.
This yields an easily computable form for the slice
\begin{equation}
\label{eq:slice}
\{g^*(y)=g, \eta(g^*(y))=\eta\}.
\end{equation}

@@

\jonathan{make the tick marks denoting the number of true groups larger -- can you not just shade the right hand part gray ? easy to do in python...}
Finally, most of the plots below show the following information. The
horizontal axis is the step index for forward stepwise. The dashed
line shows the proportion of iterations where a truly nonzero variable
was added to the model at that step. Solid verticle lines show the
middle 50\% of p-values calculated at that step, with red calculated
on a null model with no nonzero groups, blue calculated on the non-null
model, and green calculated on the non-null model from the usual
$\chi^2$ significance test.
The points in these lines are the average p-value at that step, and
the triangles above and below show the 95\% and 5\% quantiles.
